---
title: "EPR Status Variation at the ADM-2 Level"
subtitle: "Panel diagnostics and feasibility of cross-sectional analysis"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    df-print: kable
execute:
  warning: false
  message: false
---

## Overview

This document assesses whether a **time-invariant (cross-sectional) variable** is a feasible and meaningful representation of ethnic group political status for ADM-2 subnational units, using the EPR panel (2000–2022).

### What is the panel?

The panel (`adm2_epr_panel.rds`) links two datasets:

- **EPR Core**: records which ethnic group holds which political status and when. The data are structured as *status episodes* — contiguous periods during which a group holds a given status without interruption (e.g., "the Tutsi were classified as Powerless from 1990 to 2003"). Each group can have multiple episodes if its status changed over time.
- **GeoEPR**: provides the geographic polygons of each group's territorial base, also structured as episodes that can change if the group's territorial control shifted.

These are intersected against GADM 4.10 ADM-2 boundaries. The result is a **balanced panel of 48,010 subnational units × 23 years (2000–2022)**, with ten variables measuring low-power and high-power ethnic group presence in each unit — unit-relative (area and population share) and country-relative (share of country area/population).

### What do the variables measure?

| Variable | Type | Definition |
|----------|------|-----------|
| `EPR_low_bin` | Binary | Any DISCRIMINATED or POWERLESS group overlaps this ADM-2 unit |
| `EPR_high_bin` | Binary | Any MONOPOLY, DOMINANT, or SENIOR PARTNER group overlaps |
| `EPR_low_pct` | Continuous [0,1] | Share of ADM-2 *area* covered by low-power group territory |
| `EPR_high_pct` | Continuous [0,1] | Share of ADM-2 *area* covered by high-power group territory |
| `EPR_low_pop_pct` | Continuous [0,1] | Share of ADM-2 *population* inside low-power territory |
| `EPR_high_pop_pct` | Continuous [0,1] | Share of ADM-2 *population* inside high-power territory |
| `EPR_low_country_area_pct` | Continuous [0,1] | Share of *country total area* contributed by this unit's low-power territory |
| `EPR_high_country_area_pct` | Continuous [0,1] | Share of *country total area* contributed by this unit's high-power territory |
| `EPR_low_country_pop_pct` | Continuous [0,1] | Share of *country total population* contributed by this unit's low-power territory |
| `EPR_high_country_pop_pct` | Continuous [0,1] | Share of *country total population* contributed by this unit's high-power territory |

**Low-power** groups are those the EPR codes as *Discriminated* or *Powerless*. **High-power** groups hold *Monopoly*, *Dominant*, or *Senior Partner* positions. A unit can simultaneously have both (different groups in different parts of the territory).


```{r setup}
#| echo: false
library(here)
library(tidyverse)
library(kableExtra)
library(scales)
library(sf)

panel <- readRDS(here::here("output", "adm2_epr_panel.rds"))

# GADM name lookup for display in tables (country name + ADM-2 name)
gadm_names <- sf::st_read(
  here::here("source", "gadm_410_adm2_clean.gpkg"),
  quiet = TRUE
) |>
  sf::st_drop_geometry() |>
  dplyr::select(gid_adm2, iso3, country, name_adm2)
```

```{r within-unit-variation}
#| echo: false

# Collapse to one row per ADM-2 unit with within-unit variation statistics.
unit_stats <- panel |>
  group_by(gid_adm2) |>
  summarise(
    # n_distinct: 1 = never changed, 2 = changed at least once
    n_vals_low_bin   = n_distinct(EPR_low_bin),
    n_vals_high_bin  = n_distinct(EPR_high_bin),
    # abs(diff()) counts each 0→1 and 1→0 transition as one switch
    n_switches_low   = sum(abs(diff(EPR_low_bin))),
    n_switches_high  = sum(abs(diff(EPR_high_bin))),
    ever_low_bin     = any(EPR_low_bin == 1),
    always_low_bin   = all(EPR_low_bin == 1),
    ever_high_bin    = any(EPR_high_bin == 1),
    always_high_bin  = all(EPR_high_bin == 1),
    # Round before n_distinct to suppress floating-point noise from spatial joins
    n_vals_low_pct   = n_distinct(round(EPR_low_pct,  3)),
    n_vals_high_pct  = n_distinct(round(EPR_high_pct, 3)),
    sd_low_pct       = sd(EPR_low_pct),
    sd_high_pct      = sd(EPR_high_pct),
    range_low_pct    = max(EPR_low_pct) - min(EPR_low_pct),
    range_high_pct   = max(EPR_high_pct) - min(EPR_high_pct),
    sd_low_pop       = sd(EPR_low_pop_pct),
    sd_high_pop      = sd(EPR_high_pop_pct),
    range_low_pop    = max(EPR_low_pop_pct) - min(EPR_low_pop_pct),
    range_high_pop   = max(EPR_high_pop_pct) - min(EPR_high_pop_pct),
    .groups = "drop"
  ) |>
  mutate(
    changes_low_bin  = n_vals_low_bin  > 1,
    changes_high_bin = n_vals_high_bin > 1,
    changes_low_pct  = n_vals_low_pct  > 1,
    changes_high_pct = n_vals_high_pct > 1
  )

n_units <- nrow(unit_stats)
```

---

## Volatile and Transitioning Units

```{r volatile-units}
#| echo: false

# Top-30 units by within-unit SD of EPR_low_pct (the "right tail")
volatile_units <- unit_stats |>
  arrange(desc(sd_low_pct)) |>
  slice_head(n = 30) |>
  left_join(gadm_names, by = "gid_adm2") |>
  select(
    `ADM-2 ID`    = gid_adm2,
    ISO           = iso3,
    Country       = country,
    `ADM-2 Name`  = name_adm2,
    `SD (area)`   = sd_low_pct,
    `Range (area)`= range_low_pct,
    `SD (pop)`    = sd_low_pop,
    `Ever low`    = ever_low_bin,
    `Always low`  = always_low_bin,
    `Switches`    = n_switches_low
  ) |>
  mutate(across(c(`SD (area)`, `Range (area)`, `SD (pop)`), \(x) round(x, 3)))

volatile_units |>
  kbl(caption = "Top 30 ADM-2 units by within-unit SD of EPR_low_pct (most volatile)",
      align = c("l", "l", "l", "l", rep("r", 6))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) |>
  footnote(general = "'SD (area)' is the standard deviation of EPR_low_pct across 23 years for that unit. 'Switches' counts binary 0↔1 transitions. A unit can have high SD but zero switches (area share fluctuates while the binary flag stays 1).",
           general_title = "Note:")
```

```{r transient-units}
#| echo: false

# Units with transient binary exposure: ever_low = 1 but majority_low = 0
# units low-status in at least one year, but fewer than half of the 23 years
transient_units <- unit_stats |>
  filter(ever_low_bin, !always_low_bin) |>   # had at least one 0 and one 1
  left_join(
    panel |>
      group_by(gid_adm2) |>
      summarise(pct_yrs_low = mean(EPR_low_bin), .groups = "drop"),
    by = "gid_adm2"
  ) |>
  filter(pct_yrs_low < 0.5) |>              # low status in fewer than half the years
  left_join(gadm_names, by = "gid_adm2") |>
  arrange(desc(n_switches_low)) |>
  slice_head(n = 30) |>
  select(
    `ADM-2 ID`      = gid_adm2,
    ISO             = iso3,
    Country         = country,
    `ADM-2 Name`    = name_adm2,
    `% yrs low`     = pct_yrs_low,
    `Switches`      = n_switches_low,
    `SD (area)`     = sd_low_pct,
    `Range (area)`  = range_low_pct
  ) |>
  mutate(
    `% yrs low`   = percent(`% yrs low`, 0.1),
    across(c(`SD (area)`, `Range (area)`), \(x) round(x, 3))
  )

transient_units |>
  kbl(caption = "Top 30 ADM-2 units with transient low-power exposure (ever low, but in fewer than half of years), sorted by number of binary switches",
      align = c("l", "l", "l", "l", rep("r", 4))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) |>
  footnote(general = "These units are classified as 'exposed' under ever_low_bin but as 'unexposed' under majority_low_bin. They represent ambiguous cases where the choice of time-invariant construction matters.",
           general_title = "Note:")
```

---

## Part 2: Feasibility of a Time-Invariant Variable

If status is largely stable within units, a **time-invariant variable** — a single value per ADM-2 unit summarising its EPR exposure across the full period — becomes both feasible and meaningful for cross-sectional analysis.

We evaluate four candidate constructions for the binary and continuous cases.

```{r ti-candidates}
#| echo: false

ti <- panel |>
  group_by(gid_adm2) |>
  summarise(
    # "ever low": broadest — captures any exposure, even transient
    ever_low_bin     = as.integer(any(EPR_low_bin == 1)),
    # "majority years low": requires low status in ≥50% of years
    modal_low_bin    = as.integer(mean(EPR_low_bin) >= 0.5),
    # Share of years in low status (continuous version of the above)
    pct_yrs_low      = mean(EPR_low_bin),
    # Time-average of area share
    mean_low_pct     = mean(EPR_low_pct),
    # First observed value (predetermined, no post-2000 reverse causality)
    initial_low_pct  = first(EPR_low_pct),
    # Median: robust to outlier years from territorial reclassifications
    modal_low_pct    = median(EPR_low_pct),
    # Year-2000 snapshot as a predetermined anchor
    low_bin_2000     = EPR_low_bin[year == 2000],
    low_pct_2000     = EPR_low_pct[year == 2000],
    .groups = "drop"
  )
```

### 2a. Binary candidates

```{r binary-candidates-table}
#| echo: false

tibble(
  Candidate = c(
    "Ever low (any year)",
    "Majority years low (≥50%)",
    "Always low (all 23 years)",
    "Never low"
  ),
  `n units` = c(
    sum(ti$ever_low_bin),
    sum(ti$modal_low_bin),
    sum(unit_stats$always_low_bin),
    sum(!unit_stats$ever_low_bin)
  ),
  `% units` = percent(c(
    mean(ti$ever_low_bin),
    mean(ti$modal_low_bin),
    mean(unit_stats$always_low_bin),
    mean(!unit_stats$ever_low_bin)
  ), 0.1),
  `Description` = c(
    "Widest net — includes transient exposure",
    "Conservative — chronic exposure only",
    "Subset of majority: never non-low",
    "Baseline: no exposure in any year"
  )
) |>
  kbl(caption = "Binary time-invariant candidates",
      align = c("l", "r", "r", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The gap between *ever low* and *majority years low* reveals how many units had **transient** exposure — low-status groups present for some years but not the majority. These units are the key point of disagreement between the two constructions.

```{r disagreement-table}
#| echo: false

ct <- table(ever_low = ti$ever_low_bin, majority_low = ti$modal_low_bin)
n_disagree <- sum(ti$ever_low_bin != ti$modal_low_bin)

as.data.frame.matrix(ct) |>
  rownames_to_column("ever_low_bin") |>
  mutate(ever_low_bin = ifelse(ever_low_bin == "0", "Never low (0)", "Ever low (1)")) |>
  rename(`Majority non-low (0)` = `0`, `Majority low (1)` = `1`) |>
  kbl(caption = sprintf("Cross-tab: ever_low_bin × majority_low_bin. Units that disagree: %d / %d (%.1f%%)",
                        n_disagree, n_units, n_disagree / n_units * 100),
      align = c("l", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  footnote(general = "Off-diagonal cells are units classified differently by the two constructions. These are units with low-status group present in some years but fewer than half.",
           general_title = "Note:")
```

---

### 2b. Continuous candidates

```{r pct-yrs-distribution}
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Distribution of the share of years each ADM-2 unit spends in low-power status (pct_yrs_low). The two spikes at 0 and 1 correspond to permanently unexposed and permanently exposed units respectively."

ggplot(ti, aes(x = pct_yrs_low)) +
  geom_histogram(bins = 50, fill = "#66c2a5", color = "white", linewidth = 0.2) +
  scale_x_continuous(labels = percent, breaks = seq(0, 1, 0.25)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Share of years unit has EPR_low_bin = 1",
       y = "Number of ADM-2 units") +
  theme_minimal(base_size = 12)
```

The distribution is heavily bimodal: most units are either always exposed or never exposed. The middle — units that change status partway through the period — is thin. This is encouraging for a time-invariant variable: **the binary classification will be stable for the large majority of units**.

```{r continuous-candidates-summary}
#| echo: false

tibble(
  Candidate = c("Mean area share (mean_low_pct)",
                "Initial value / year 2000 (initial_low_pct)",
                "Median area share (modal_low_pct)"),
  Mean   = round(c(mean(ti$mean_low_pct), mean(ti$initial_low_pct), mean(ti$modal_low_pct)), 4),
  Median = round(c(median(ti$mean_low_pct), median(ti$initial_low_pct), median(ti$modal_low_pct)), 4),
  SD     = round(c(sd(ti$mean_low_pct), sd(ti$initial_low_pct), sd(ti$modal_low_pct)), 4),
  `% zero` = percent(c(mean(ti$mean_low_pct == 0),
                        mean(ti$initial_low_pct == 0),
                        mean(ti$modal_low_pct == 0)), 0.1)
) |>
  kbl(caption = "Summary statistics for continuous time-invariant candidates",
      align = c("l", rep("r", 4))) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  footnote(general = "A high % zero reflects units with no low-power group overlap in any year (or only in the reference year for the initial value). Between-unit SD is the key figure for cross-section identification: higher SD = more variation to exploit.",
           general_title = "Note:")
```

```{r correlation-heatmap}
#| echo: false
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Correlation matrix of time-invariant candidate variables. High correlations indicate the choice of construction makes little difference to unit rankings."

cor_mat <- ti |>
  select(
    `Ever low\n(binary)`      = ever_low_bin,
    `Majority low\n(binary)`  = modal_low_bin,
    `% years low\n(continuous)` = pct_yrs_low,
    `Mean area\nshare`         = mean_low_pct,
    `Initial area\nshare`      = initial_low_pct,
    `Year-2000\narea share`    = low_pct_2000
  ) |>
  cor(use = "complete.obs")

cor_mat |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "cor") |>
  mutate(var1 = factor(var1, levels = colnames(cor_mat)),
         var2 = factor(var2, levels = colnames(cor_mat))) |>
  ggplot(aes(x = var2, y = fct_rev(var1), fill = cor)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(cor, 2)), size = 3.5) +
  scale_fill_gradient2(low = "#f7fbff", mid = "#6baed6", high = "#08306b",
                       midpoint = 0.5, limits = c(0, 1),
                       name = "Correlation") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "right")
```

**How to read the correlation matrix:** Each cell shows the Pearson correlation between two candidate variables computed across all 48,010 ADM-2 units. A value of 1.00 on the diagonal is trivial (a variable is perfectly correlated with itself). The off-diagonal values are what matter.

A **high correlation** (e.g., 0.90+) between two candidates means that units rank similarly on both measures — knowing one tells you almost everything about the other. In that case, the choice between them is largely cosmetic: results in a regression will be nearly identical regardless of which you use.

A **low correlation** would mean the two candidates assign very different exposure values to some units — choosing one vs. the other would change which units are classified as "exposed" and could meaningfully alter regression results.

**What to look for here:**

- *Ever-exposed vs. majority-exposed*: the correlation between `ever_low_bin` and `majority_low_bin` tells you how often the two disagree. The off-diagonal disagreements are units low-status in some years but fewer than half — the transient-exposure cases identified in the Volatile and Transitioning Units section above.
- *Binary vs. continuous*: these correlations are naturally lower because a binary variable loses information about *how much* of the unit is covered. Even so, a high correlation means that units with large area shares tend to be the ones flagged as low-power.
- *Long-run mean vs. baseline snapshot*: correlations among `mean_low_pct`, `initial_low_pct`, and `low_pct_2000` tell you whether the 23-year average is well-approximated by a single-year snapshot. If yes, using the year-2000 value as a predetermined variable is defensible without much loss of information.

---

## Summary

```{r variation-summary}
#| echo: false

tibble(
  Finding = c(
    "Share of ADM-2 units with stable EPR_low_bin",
    "Share of ADM-2 units with stable EPR_high_bin",
    "Median within-unit SD of EPR_low_pct",
    "Units permanently unexposed (EPR_low_bin always 0)",
    "Units permanently exposed (EPR_low_bin always 1)",
    "Units with transient exposure (ever but not majority low)",
    "Correlation: ever_low_bin vs mean_low_pct"
  ),
  Value = c(
    percent(mean(!unit_stats$changes_low_bin), 0.1),
    percent(mean(!unit_stats$changes_high_bin), 0.1),
    formatC(median(unit_stats$sd_low_pct), digits = 4, format = "f"),
    comma(sum(!unit_stats$ever_low_bin)),
    comma(sum(unit_stats$always_low_bin)),
    comma(n_disagree),
    formatC(cor(ti$ever_low_bin, ti$mean_low_pct, use = "complete.obs"), digits = 3, format = "f")
  )
) |>
  kbl(caption = "Key diagnostics at a glance", align = c("l", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Take-aways for the co-author:**

1. **Most units are time-stable in the binary sense.** The large majority never cross the low-power/high-power boundary. This limits the scope for panel identification via within-unit variation, but makes cross-sectional analysis well-motivated.

2. **The continuous variable moves more than the binary, but not by much.** Median within-unit SD is close to zero. Variation in the continuous measure comes largely from the tail of volatile units.

3. **A time-invariant variable is feasible.** Candidates agree closely with one another (high correlations). The *ever-low* binary and the *mean area share* are natural choices: the former for clean interpretation, the latter for a richer continuous exposure measure.

4. **Transient exposure is the main ambiguity.** Units that were low-status for some years but not the majority are classified differently by *ever-low* vs *majority-low*. The share of such units and whether they differ systematically from chronic-exposure units is worth examining in the substantive analysis.

---

```{r session-info}
#| echo: false
#| output: false
sessionInfo()
```

*Generated `r format(Sys.time(), "%Y-%m-%d %H:%M")`*
