---
title: "ADM-2 × Year EPR Political Status Panel"
format: html
---

```{r setup}
# No stochastic operations in this script; parallel workers use future.seed = TRUE.
if (!require("pacman")) install.packages("pacman", repos = "https://cran.r-project.org")

pacman::p_load(
  here,
  tidyverse,
  janitor,
  sf,
  terra,
  exactextractr,
  units,
  future,
  future.apply
)

```

```{r constants}
# Disable S2 spherical geometry engine — use planar GEOS instead.
# S2 is stricter about self-intersecting rings (common in global datasets like GeoEPR).
sf::sf_use_s2(FALSE)

# Parallelism: multisession works on all platforms (including Windows).
# Detect cores; leave one free for the OS.
n_cores <- max(1L, parallel::detectCores() - 1L)
plan(multisession, workers = n_cores)
cat("Parallel workers:", n_cores, "\n")

YEAR_START     <- 2000L
YEAR_END       <- 2023L
RC1_CHECK_YEAR <- 2010L

# ── Data paths — relative to project root via here::here() ───────────────────
GADM_PATH     <- here::here("EPR/source/gadm_410_adm2_clean.gpkg")
EPR_PATH      <- here::here("EPR/source/EPR/EPR-2023.csv")
GEOEPR_PATH   <- here::here("EPR/source/EPR/GeoEPR-2023.geojson")
WORLDPOP_PATH <- here::here("EPR/source/WorldPop/ppp_2000_1km_Aggregated.tif")

low_status  <- c("DISCRIMINATED", "POWERLESS")
high_status <- c("MONOPOLY", "DOMINANT", "SENIOR PARTNER")

mollweide_crs <- st_crs("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# ── Caching: skip expensive recomputation on re-render ───────────────────────
# Set FORCE_RECOMPUTE = TRUE to bust the cache and rerun from scratch.
FORCE_RECOMPUTE <- FALSE
CACHE_LINKS <- here::here("EPR", "output", "cache_adm2_spell_links.rds")
CACHE_ISECT <- here::here("EPR", "output", "cache_spell_intersections_sf.rds")
CACHE_POP   <- here::here("EPR", "output", "cache_pop_overlap.rds")

# ── Expensive robustness check flag ─────────────────────────────────────────
# RC1 reruns two full st_join() passes (WGS84 + Mollweide). Already confirmed
# stable (4 ADM-2 units differ). Set TRUE only to re-validate after data changes.
RUN_EXPENSIVE_RCS <- FALSE
```

## Step 1: Load GADM ADM-2

```{r load-gadm}
adm2 <- st_read(GADM_PATH, quiet = TRUE)
cat("GADM ADM-2:", nrow(adm2), "units\n")
cat("CRS:", st_crs(adm2)$input, "\n")
```

## Step 2: Load EPR + GeoEPR

```{r load-epr}
epr_data    <- readr::read_csv(EPR_PATH,    show_col_types = FALSE)
geoepr_data <- sf::st_read(GEOEPR_PATH, quiet = TRUE)

cat("EPR rows:", nrow(epr_data), "| GeoEPR rows:", nrow(geoepr_data), "\n")
cat("EPR status values:\n"); print(table(epr_data$status))
cat("GeoEPR CRS:", st_crs(geoepr_data)$input, "\n")
```

```{r expand-epr-yearly}
# Expand EPR Core to yearly panel — this gives us status per gwgroupid × year.
# GeoEPR does NOT need yearly expansion; spell polygons are used directly for
# the spatial join (see Step 4).
expand_yearly <- function(df, year_start, year_end) {
  df |>
    dplyr::filter(to >= year_start, from <= year_end) |>
    dplyr::mutate(
      year_from = pmax(as.integer(from), year_start),
      year_to   = pmin(as.integer(to),   year_end),
      year      = purrr::map2(year_from, year_to, seq)
    ) |>
    tidyr::unnest(year) |>
    dplyr::select(-year_from, -year_to)
}

epr_yearly <- expand_yearly(epr_data, YEAR_START, YEAR_END)
cat("EPR yearly rows:", nrow(epr_yearly),
    "| years:", paste(range(epr_yearly$year), collapse = "-"), "\n")

# Coverage diagnostic: EPR groups without GeoEPR geometry
epr_ids    <- distinct(epr_data,    gwgroupid)
geoepr_ids <- geoepr_data |> st_drop_geometry() |> distinct(gwgroupid)
epr_only   <- anti_join(epr_ids, geoepr_ids, by = "gwgroupid")
cat("EPR groups without GeoEPR geometry (lost from overlay):", nrow(epr_only), "\n")
if (nrow(epr_only) > 0) {
  epr_data |>
    filter(gwgroupid %in% epr_only$gwgroupid) |>
    select(any_of(c("gwgroupid", "group", "statename", "status", "from", "to"))) |>
    distinct() |> arrange(gwgroupid) |> print(n = 10)
}
```

## Coverage Audit — Phase 1: Characterize Missing Groups

```{r coverage-phase1}
cat(sprintf("=== COVERAGE AUDIT — Phase 1: %d EPR groups without GeoEPR polygon ===\n\n",
            nrow(epr_only)))

# Full profile of missing groups using all available EPR columns
epr_missing <- epr_data |>
  filter(gwgroupid %in% epr_only$gwgroupid) |>
  select(any_of(c("gwgroupid", "group", "statename", "status",
                  "from", "to", "size", "reg_aut", "umbrella"))) |>
  distinct()

# Unique groups (not spells)
missing_groups <- epr_missing |>
  distinct(gwgroupid, group, statename)
cat("Unique groups without geometry:", nrow(missing_groups), "\n\n")

# Summary by EPR status
cat("--- By EPR status (spell-level) ---\n")
epr_missing |>
  count(status, sort = TRUE) |>
  print()

# Summary by country
cat("\n--- By country (top 20) ---\n")
missing_groups |>
  count(statename, sort = TRUE) |>
  print(n = 20)

# Bias assessment: do any excluded groups ever hold DISCRIMINATED / POWERLESS status?
cat("\n--- Bias assessment ---\n")
bias_check <- epr_missing |>
  group_by(gwgroupid, group, statename) |>
  summarise(
    has_low_status  = any(status %in% low_status),
    has_high_status = any(status %in% high_status),
    .groups = "drop"
  )

cat("Groups with ≥1 spell in {DISCRIMINATED, POWERLESS}:",
    sum(bias_check$has_low_status), "/", nrow(bias_check),
    sprintf("(%.0f%%)\n", mean(bias_check$has_low_status) * 100))
cat("Groups with ≥1 spell in {MONOPOLY, DOMINANT, SENIOR PARTNER}:",
    sum(bias_check$has_high_status), "/", nrow(bias_check),
    sprintf("(%.0f%%)\n", mean(bias_check$has_high_status) * 100))

cat("\nIf the 153 are overwhelmingly dispersed (non-regional) groups,",
    "their exclusion is expected by GeoEPR design — not a coverage error.\n")
cat("If some are regionally based with low status, their exclusion could",
    "downward-bias EPR_low_bin in affected countries.\n")

# Save full group list for inspection
csv_path <- here::here("EPR", "output", "epr_groups_no_geometry.csv")
epr_missing |>
  arrange(gwgroupid, from) |>
  write_csv(csv_path)
cat("\nSaved:", csv_path, "\n")
```

## Step 3: CRS Alignment and Geometry Validation

```{r crs-align}
target_crs <- st_crs(4326)

if (st_crs(adm2) != target_crs) {
  cat("Reprojecting GADM to WGS84...\n")
  adm2 <- st_transform(adm2, target_crs)
}
if (st_crs(geoepr_data) != target_crs) {
  cat("Reprojecting GeoEPR to WGS84...\n")
  geoepr_data <- st_transform(geoepr_data, target_crs)
}

n_inv_adm2   <- sum(!st_is_valid(adm2))
n_inv_geoepr <- sum(!st_is_valid(geoepr_data))
cat("Invalid GADM geometries:", n_inv_adm2, "\n")
cat("Invalid GeoEPR geometries:", n_inv_geoepr, "\n")
if (n_inv_adm2   > 0) adm2        <- st_make_valid(adm2)
if (n_inv_geoepr > 0) geoepr_data <- st_make_valid(geoepr_data)
```

## Coverage Audit — Phase 1b: GeoEPR Geometry Changes Across Spells

```{r coverage-phase1b}
cat("=== COVERAGE AUDIT — Phase 1b: Do GeoEPR groups change geometry across spells? ===\n\n")

# Compute area + centroid as a geometry fingerprint per spell
geoepr_fingerprint <- geoepr_data |>
  mutate(
    spell_area_km2 = suppressWarnings(as.numeric(st_area(geometry))) / 1e6,
    cx = suppressWarnings(st_coordinates(st_centroid(geometry))[, 1]),
    cy = suppressWarnings(st_coordinates(st_centroid(geometry))[, 2])
  ) |>
  st_drop_geometry() |>
  select(gwgroupid, from, to, spell_area_km2, cx, cy)

# Per group: count distinct area + centroid combinations across spells
geom_stability <- geoepr_fingerprint |>
  group_by(gwgroupid) |>
  summarise(
    n_spells           = n(),
    n_distinct_areas   = n_distinct(round(spell_area_km2, 1)),
    n_distinct_centers = n_distinct(paste(round(cx, 3), round(cy, 3))),
    .groups = "drop"
  ) |>
  mutate(geometry_changes = n_distinct_areas > 1 | n_distinct_centers > 1)

cat("Total GeoEPR groups:                     ", nrow(geom_stability), "\n")
cat("  Single-spell groups:                   ", sum(geom_stability$n_spells == 1), "\n")
cat("  Multi-spell groups:                    ", sum(geom_stability$n_spells > 1), "\n")
cat("  Multi-spell with CHANGING geometry:    ",
    sum(geom_stability$geometry_changes), "\n")
cat("  Multi-spell with STABLE geometry:      ",
    sum(geom_stability$n_spells > 1 & !geom_stability$geometry_changes), "\n")

# Are geometry changes handled by the current script?
cat("\n--- Assessment ---\n")
cat("The script joins at spell level (geoepr_spells retains from/to per polygon).\n")
cat("Each spell's geometry is used independently in the st_join, so groups with\n")
cat("changing geometries ARE captured correctly — their ADM-2 intersections can\n")
cat("differ across time periods, and the tabular year expansion (Step 5) propagates\n")
cat("each spell's ADM-2 links only within its [from, to] interval.\n")

if (sum(geom_stability$geometry_changes) > 0) {
  cat("\nSample of groups with geometry changes across spells:\n")
  geom_stability |>
    filter(geometry_changes) |>
    arrange(desc(n_spells)) |>
    print(n = 10)
}
```

## Step 4: Precomputed Spatial Join (runs once)

```{r precompute-links}
# ── Strategy ─────────────────────────────────────────────────────────────────
# Old approach: st_union + st_intersects called 48× (24 years × 2 statuses).
# New approach: one st_join against the ~1684 raw GeoEPR spell polygons.
#   The ADM-2 × GeoEPR spatial relationship is fixed within each spell —
#   no need to recompute per year. All yearly logic is then tabular.

if (!FORCE_RECOMPUTE && file.exists(CACHE_LINKS)) {
  cat("Loading cached spell links...\n")
  adm2_spell_links <- readRDS(CACHE_LINKS)
} else {
  cat("Precomputing ADM-2 × GeoEPR spell links...\n")

  # Keep only the columns we need from GeoEPR spells
  geoepr_spells <- geoepr_data |> select(gwgroupid, from, to)

  # st_join: for each ADM-2 polygon, find all GeoEPR spell polygons it intersects.
  # left = FALSE drops ADM-2 units with zero EPR coverage (added back as 0 later).
  adm2_spell_links <- st_join(
    adm2 |> select(gid_adm2),
    geoepr_spells,
    join = st_intersects,
    left = FALSE
  ) |>
    st_drop_geometry() |>
    as_tibble()

  saveRDS(adm2_spell_links, CACHE_LINKS)
  cat("Cached:", CACHE_LINKS, "\n")
}

cat("Spell links found:", nrow(adm2_spell_links),
    "(ADM-2 × GeoEPR spell intersections)\n")
cat("Unique ADM-2 units with any EPR overlap:",
    n_distinct(adm2_spell_links$gid_adm2), "\n")
cat("Unique GeoEPR spells matched:",
    n_distinct(paste(adm2_spell_links$gwgroupid, adm2_spell_links$from)), "\n")
```

## Step 4b: Intersection Areas (Mollweide, runs once)

```{r compute-areas}
# Compute overlap_pct = int_area / adm2_area for every (ADM-2, GeoEPR spell) pair.
# Used in Step 5 to build continuous EPR variables AND reused by RC2 for threshold checks.
# Running in the main pipeline avoids duplicating the heavy pmap_dfr loop in RC2.

# ── Project to Mollweide (equal-area) — always needed ────────────────────────
adm2_moll <- adm2 |> sf::st_transform(mollweide_crs)
adm2_moll$adm2_area_m2 <- as.numeric(sf::st_area(adm2_moll))
adm2_moll <- adm2_moll |> dplyr::select(gid_adm2, adm2_area_m2)

geoepr_moll <- geoepr_data |>
  sf::st_transform(mollweide_crs) |>
  dplyr::select(gwgroupid, from, to)

if (!FORCE_RECOMPUTE && file.exists(CACHE_ISECT)) {
  cat("Loading cached spell intersections...\n")
  spell_intersections_sf <- readRDS(CACHE_ISECT)
} else {
  cat("Computing intersection areas in Mollweide (vectorized, runs once)...\n")

  # Subset adm2_moll to only units with any EPR overlap — reduces size before join.
  adm2_subset <- adm2_moll |>
    dplyr::filter(gid_adm2 %in% adm2_spell_links$gid_adm2) |>
    sf::st_make_valid()

  geoepr_moll_valid <- geoepr_moll |> sf::st_make_valid()

  # Single vectorized intersection — replaces the serial pmap() loop (~740 iters).
  # GEOS uses an STR-tree spatial index: O((n+m) log(n+m)) vs 740 separate calls.
  spell_intersections_sf <- suppressWarnings(
    sf::st_intersection(adm2_subset, geoepr_moll_valid)
  )
  # Remove empty geometries; bracket subsetting avoids assuming a geometry column name.
  spell_intersections_sf <- spell_intersections_sf[!sf::st_is_empty(spell_intersections_sf), ]
  # Extract polygon parts from any GEOMETRYCOLLECTION results.
  spell_intersections_sf <- sf::st_collection_extract(spell_intersections_sf, "POLYGON")
  spell_intersections_sf <- dplyr::mutate(spell_intersections_sf,
      int_area_m2 = as.numeric(sf::st_area(spell_intersections_sf)),
      overlap_pct = int_area_m2 / adm2_area_m2
    ) |>
    dplyr::rename(spell_from = from, spell_to = to) |>
    dplyr::select(gid_adm2, gwgroupid, spell_from, spell_to, overlap_pct)
  # Geometry retained intentionally — needed by Step 4c for population extraction.

  saveRDS(spell_intersections_sf, CACHE_ISECT)
  cat("Cached:", CACHE_ISECT, "\n")
}

# Split into two objects with distinct purposes:
#   spell_intersections      — tabular (no geometry), used for area-based EPR_*_pct in Step 5
#   spell_intersection_geoms — sf with geometry, used for pop zonal extraction in Step 4c
# Splitting avoids carrying large geometry through the tabular pipeline unnecessarily.
spell_intersections      <- spell_intersections_sf |> sf::st_drop_geometry()
spell_intersection_geoms <- spell_intersections_sf  # keep sf class
rm(spell_intersections_sf)                          # free the intermediate object

cat("Spell intersections computed:", nrow(spell_intersections), "\n")
```

## Step 4c: Population-Weighted Overlap (WorldPop, year 2000)

```{r compute-pop-weights}
# ── Why WorldPop year 2000? ───────────────────────────────────────────────────
# We anchor population weights to year 2000 (the panel start) for two reasons:
#   1. Avoids endogeneity: population shifts over 2000–2023 may partly respond to
#      political exclusion itself (e.g., displacement, migration), so using a
#      fixed baseline prevents the weight from being contaminated by the outcome.
#   2. Consistency: the political-salience variable applies to the full panel;
#      using a fixed population grid makes the denominator stable across years,
#      so changes in EPR_*_pop_pct reflect changes in political status, not
#      changes in population distribution.
# Source: WorldPop "Estimated total number of people per grid-cell" (2000),
# 30 arc-second resolution (~1 km at equator), WGS84, units = people per pixel.

if (!FORCE_RECOMPUTE && file.exists(CACHE_POP)) {
  cat("Loading cached population-weighted overlap...\n")
  pop_overlap <- readRDS(CACHE_POP)
  rm(spell_intersection_geoms); gc()
} else {
  cat("Loading WorldPop year 2000 raster...\n")
  pop_rast <- terra::rast(WORLDPOP_PATH)
  cat("Loaded:", basename(WORLDPOP_PATH),
      "| CRS:", terra::crs(pop_rast, describe = TRUE)$name,
      "| Resolution:", paste(round(terra::res(pop_rast), 5), collapse = " x "), "deg\n")

  # ── Reproject ADM-2 to raster CRS for exact_extract ────────────────────────
  # exact_extract requires vector and raster in the same CRS.
  # WorldPop is WGS84; adm2 is also WGS84 — the transform is a no-op but kept
  # explicit so the code is robust if the raster CRS ever changes.
  adm2_for_pop <- adm2 |> sf::st_transform(terra::crs(pop_rast))

  # ── Denominator: total population per ADM-2 unit ───────────────────────────
  # exact_extract sums raster cell values weighted by fractional cell coverage,
  # giving correct totals even for cells that straddle polygon boundaries.
  cat("Extracting ADM-2 total population (denominator)...\n")
  adm2_for_pop$adm2_pop <- exactextractr::exact_extract(
    pop_rast, adm2_for_pop, fun = "sum", progress = TRUE
  )

  # ── Diagnose zero-population ADM-2 units ───────────────────────────────────
  # Zero-pop units get pop_overlap_pct = 0 (division by zero prevention).
  # We report them explicitly so the user can assess whether any are substantively
  # important (e.g., small islands with EPR groups should appear as NA, not zero).
  zero_pop <- adm2_for_pop |>
    sf::st_drop_geometry() |>
    dplyr::filter(adm2_pop == 0) |>
    dplyr::select(dplyr::any_of(c("gid_adm2", "country", "iso3", "NAME_0", "GID_0")))
  cat("ADM-2 units with zero WorldPop population (year 2000):", nrow(zero_pop), "\n")
  if (nrow(zero_pop) > 0 && all(c("iso3", "country") %in% names(zero_pop))) {
    zero_pop_country <- zero_pop |>
      dplyr::count(iso3, country, name = "n_units") |>
      dplyr::arrange(dplyr::desc(n_units))
    print(head(zero_pop_country, 20))
  }

  # ── Reproject intersection polygons to raster CRS ──────────────────────────
  # spell_intersection_geoms is in Mollweide (from Step 4b); WorldPop is WGS84.
  spell_intersection_geoms <- spell_intersection_geoms |>
    sf::st_transform(terra::crs(pop_rast))

  # ── Drop non-polygon geometries ─────────────────────────────────────────────
  # st_intersection() can return GEOMETRYCOLLECTION, LINESTRING, or POINT when
  # two polygons share only an edge or corner. exact_extract() requires strictly
  # polygonal features, so we extract polygon parts and discard degenerate results.
  n_before <- nrow(spell_intersection_geoms)
  # st_collection_extract("POLYGON") decomposes GEOMETRYCOLLECTION results into
  # pure polygon features, discarding any degenerate point/line components that
  # arise when two polygons share only an edge or corner.
  spell_intersection_geoms <- sf::st_collection_extract(spell_intersection_geoms, "POLYGON")
  cat("Non-polygon geometries dropped:",
      n_before - nrow(spell_intersection_geoms),
      "of", n_before, "intersection rows\n")

  # ── Numerator: population within each (ADM-2 × GeoEPR spell) intersection ──
  # Each row of spell_intersection_geoms is the polygon of the overlap between one
  # ADM-2 unit and one GeoEPR group spell. Summing population within that polygon
  # gives the number of people in the ADM-2 who live inside the group's territory.
  cat("Extracting intersection population (numerator)...\n")
  spell_intersection_geoms$int_pop <- exactextractr::exact_extract(
    pop_rast, spell_intersection_geoms, fun = "sum", progress = TRUE
  )

  # ── Compute population overlap fraction ────────────────────────────────────
  # pop_overlap_pct = people in (ADM-2 ∩ GeoEPR group) / total people in ADM-2.
  # This is the population-weighted analogue of overlap_pct (area-based).
  # Zero-pop units are set to 0, not NA, to keep the panel fully non-missing.
  adm2_pop_lookup <- adm2_for_pop |>
    sf::st_drop_geometry() |>
    dplyr::select(gid_adm2, adm2_pop)

  pop_overlap <- spell_intersection_geoms |>
    sf::st_drop_geometry() |>
    dplyr::select(gid_adm2, gwgroupid, spell_from, spell_to, int_pop) |>
    dplyr::left_join(adm2_pop_lookup, by = "gid_adm2") |>
    dplyr::mutate(
      pop_overlap_pct = dplyr::if_else(adm2_pop > 0, int_pop / adm2_pop, 0)
    )

  saveRDS(pop_overlap, CACHE_POP)
  cat("Cached:", CACHE_POP, "\n")

  # ── Free large objects ──────────────────────────────────────────────────────
  # spell_intersection_geoms (large sf) and the raster are no longer needed.
  # Free them before building the panel to keep peak memory manageable.
  rm(spell_intersection_geoms, pop_rast, adm2_for_pop)
  gc()
}

cat("pop_overlap rows:", nrow(pop_overlap),
    "| range pop_overlap_pct: [",
    round(min(pop_overlap$pop_overlap_pct), 4), ",",
    round(max(pop_overlap$pop_overlap_pct), 4), "]\n")
```

## Step 5: Build Balanced Panel

```{r build-panel}
# ── Expand spell links to yearly grain ───────────────────────────────────────
adm2_group_years <- expand_yearly(adm2_spell_links, YEAR_START, YEAR_END) |>
  select(gid_adm2, gwgroupid, year)

# ── Join EPR status and aggregate — binary flags ──────────────────────────────
flags_df <- adm2_group_years |>
  left_join(
    epr_yearly |> select(gwgroupid, year, status),
    by = c("gwgroupid", "year")
  ) |>
  group_by(gid_adm2, year) |>
  summarise(
    EPR_low_bin  = as.integer(any(status %in% low_status,  na.rm = TRUE)),
    EPR_high_bin = as.integer(any(status %in% high_status, na.rm = TRUE)),
    .groups = "drop"
  )

# ── Continuous variables: % of ADM-2 area covered by low/high status groups ──
# overlap_pct = int_area / adm2_area (from Step 4b).
# Summed across same-status groups, capped at 1.
continuous_df <- expand_yearly(
    spell_intersections |> rename(from = spell_from, to = spell_to),
    YEAR_START, YEAR_END
  ) |>
  select(gid_adm2, gwgroupid, year, overlap_pct) |>
  left_join(epr_yearly |> select(gwgroupid, year, status),
            by = c("gwgroupid", "year")) |>
  group_by(gid_adm2, year) |>
  summarise(
    EPR_low_pct  = min(sum(overlap_pct[status %in% low_status],  na.rm = TRUE), 1),
    EPR_high_pct = min(sum(overlap_pct[status %in% high_status], na.rm = TRUE), 1),
    .groups = "drop"
  )

# ── Population-weighted continuous variables ──────────────────────────────────
# Mirrors the area-based continuous_df above, but uses pop_overlap_pct instead of
# overlap_pct as the weight. This measures the share of each ADM-2's *population*
# living inside low/high-power group territories, rather than the area share.
# Population weighting matters where groups are geographically concentrated in
# dense urban areas (large pop share, small area share) or remote rural regions
# (small pop share, large area share).
pop_continuous_df <- expand_yearly(
    pop_overlap |> rename(from = spell_from, to = spell_to),
    YEAR_START, YEAR_END
  ) |>
  select(gid_adm2, gwgroupid, year, pop_overlap_pct) |>
  left_join(epr_yearly |> select(gwgroupid, year, status),
            by = c("gwgroupid", "year")) |>
  group_by(gid_adm2, year) |>
  summarise(
    # Sum population shares across groups of the same status; cap at 1.
    # Cap is needed because multiple groups can overlap the same ADM-2 population.
    EPR_low_pop_pct  = min(sum(pop_overlap_pct[status %in% low_status],  na.rm = TRUE), 1),
    EPR_high_pop_pct = min(sum(pop_overlap_pct[status %in% high_status], na.rm = TRUE), 1),
    .groups = "drop"
  )

# ── Balanced panel ────────────────────────────────────────────────────────────
panel <- expand_grid(
  gid_adm2 = unique(adm2$gid_adm2),
  year     = YEAR_START:YEAR_END
) |>
  left_join(flags_df,          by = c("gid_adm2", "year")) |>
  left_join(continuous_df,     by = c("gid_adm2", "year")) |>
  left_join(pop_continuous_df, by = c("gid_adm2", "year")) |>
  mutate(
    EPR_low_bin      = replace_na(EPR_low_bin,      0L),
    EPR_high_bin     = replace_na(EPR_high_bin,     0L),
    EPR_low_pct      = replace_na(EPR_low_pct,      0),
    EPR_high_pct     = replace_na(EPR_high_pct,     0),
    EPR_low_pop_pct  = replace_na(EPR_low_pop_pct,  0),
    EPR_high_pop_pct = replace_na(EPR_high_pop_pct, 0)
  )

# Balance check
n_units <- n_distinct(panel$gid_adm2)
n_years <- n_distinct(panel$year)
stopifnot("Panel is not balanced" = nrow(panel) == n_units * n_years)
cat("PASS: Balanced panel —", n_units, "ADM-2 units ×",
    n_years, "years =", nrow(panel), "rows\n")
cat("Columns:", paste(names(panel), collapse = ", "), "\n")
```

## Step 6: Summary Statistics

```{r summary-by-year}
summary_year <- panel |>
  group_by(year) |>
  summarise(
    n_units  = n(),
    n_low    = sum(EPR_low_bin),
    n_high   = sum(EPR_high_bin),
    pct_low  = mean(EPR_low_bin) * 100,
    pct_high = mean(EPR_high_bin) * 100,
    .groups = "drop"
  )
cat("=== EPR flag coverage by year ===\n")
print(summary_year, n = Inf)
```

```{r cross-tab}
cat("\n=== ADM-2 × year cross-tabulation (pooled) ===\n")
cat("Both low and high:", sum(panel$EPR_low_bin == 1 & panel$EPR_high_bin == 1), "\n")
cat("Low only:         ", sum(panel$EPR_low_bin == 1 & panel$EPR_high_bin == 0), "\n")
cat("High only:        ", sum(panel$EPR_low_bin == 0 & panel$EPR_high_bin == 1), "\n")
cat("Neither:          ", sum(panel$EPR_low_bin == 0 & panel$EPR_high_bin == 0), "\n")

cat("\n=== Continuous variables: EPR_low_pct / EPR_high_pct ===\n")
cat("EPR_low_pct  — mean:", round(mean(panel$EPR_low_pct),  4),
    "| median:", round(median(panel$EPR_low_pct),  4),
    "| sd:", round(sd(panel$EPR_low_pct),  4),
    "| % zeros:", round(mean(panel$EPR_low_pct  == 0) * 100, 1), "%\n")
cat("EPR_high_pct — mean:", round(mean(panel$EPR_high_pct), 4),
    "| median:", round(median(panel$EPR_high_pct), 4),
    "| sd:", round(sd(panel$EPR_high_pct), 4),
    "| % zeros:", round(mean(panel$EPR_high_pct == 0) * 100, 1), "%\n")

# Consistency check: binary == 1 iff continuous > 0
n_inconsistent_low  <- sum((panel$EPR_low_bin  == 1) != (panel$EPR_low_pct  > 0))
n_inconsistent_high <- sum((panel$EPR_high_bin == 1) != (panel$EPR_high_pct > 0))
cat("Consistency (bin==1 ↔ pct>0) — low inconsistencies:", n_inconsistent_low,
    "| high:", n_inconsistent_high, "\n")

cat("\n=== Population-weighted variables: EPR_low_pop_pct / EPR_high_pop_pct ===\n")
cat("EPR_low_pop_pct  — mean:", round(mean(panel$EPR_low_pop_pct),  4),
    "| median:", round(median(panel$EPR_low_pop_pct),  4),
    "| sd:", round(sd(panel$EPR_low_pop_pct),  4),
    "| % zeros:", round(mean(panel$EPR_low_pop_pct  == 0) * 100, 1), "%\n")
cat("EPR_high_pop_pct — mean:", round(mean(panel$EPR_high_pop_pct), 4),
    "| median:", round(median(panel$EPR_high_pop_pct), 4),
    "| sd:", round(sd(panel$EPR_high_pop_pct), 4),
    "| % zeros:", round(mean(panel$EPR_high_pop_pct == 0) * 100, 1), "%\n")

# Correlation between area-based and population-based measures.
# High correlation (> 0.7) confirms both capture similar underlying geography.
# Divergence reveals areas where population is concentrated in group territories
# that cover a small land area (e.g., urban ethnic enclaves), or vice versa.
cat("\n=== Area vs. population-weighted correlation ===\n")
cat("Cor(EPR_low_pct,  EPR_low_pop_pct): ",
    round(cor(panel$EPR_low_pct,  panel$EPR_low_pop_pct),  3), "\n")
cat("Cor(EPR_high_pct, EPR_high_pop_pct):",
    round(cor(panel$EPR_high_pct, panel$EPR_high_pop_pct), 3), "\n")

# Consistency check for pop-weighted variables: same logic as area-based.
# If binary == 1 but pop_pct == 0, it means no one lives in the intersection
# (uninhabited territory with a GeoEPR group polygon but zero GHSL population).
# These are expected but should be few.
n_incons_pop_low  <- sum((panel$EPR_low_bin  == 1) & (panel$EPR_low_pop_pct  == 0))
n_incons_pop_high <- sum((panel$EPR_high_bin == 1) & (panel$EPR_high_pop_pct == 0))
cat("Binary=1 but pop_pct=0 — low:", n_incons_pop_low,
    "| high:", n_incons_pop_high,
    "(expected: uninhabited group territories)\n")
```

```{r plot-time-trend}
summary_year |>
  select(year, pct_low, pct_high) |>
  pivot_longer(c(pct_low, pct_high), names_to = "category", values_to = "pct") |>
  mutate(category = case_match(category,
    "pct_low"  ~ "Low power (Discriminated/Powerless)",
    "pct_high" ~ "High power (Monopoly/Dominant/Senior Partner)"
  )) |>
  ggplot(aes(x = year, y = pct, color = category)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.5) +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Share of ADM-2 units overlapping EPR group territories",
       x = "Year", y = "% of ADM-2 units", color = NULL) +
  theme_bw() + theme(legend.position = "bottom")
```

```{r map-sample-year}
sample_year <- 2010L
adm2 |>
  select(gid_adm2) |>
  left_join(panel |> filter(year == sample_year), by = "gid_adm2") |>
  mutate(
    epr_category = case_when(
      EPR_low_bin == 1 & EPR_high_bin == 1 ~ "Both",
      EPR_low_bin == 1                      ~ "Low power only",
      EPR_high_bin == 1                     ~ "High power only",
      TRUE                                  ~ "Neither"
    ) |> factor(levels = c("Neither", "Low power only", "High power only", "Both"))
  ) |>
  ggplot() +
  geom_sf(aes(fill = epr_category), color = NA) +
  scale_fill_manual(
    values = c("Neither" = "grey90", "Low power only" = "steelblue",
               "High power only" = "firebrick", "Both" = "purple4"),
    na.value = "grey80"
  ) +
  labs(title = paste("EPR political status overlay —", sample_year),
       subtitle = "ADM-2 level", fill = NULL) +
  theme_void() + theme(legend.position = "bottom")
```

## Step 7: Robustness Checks

### RC1: CRS sensitivity

```{r robustness-crs}
# Re-run the precomputed spatial join in WGS84 and Mollweide in parallel.
# Compare binary flags for a spot-check year to assess CRS sensitivity.
# Gated by RUN_EXPENSIVE_RCS (default FALSE) — already confirmed: 4 units differ.

if (!RUN_EXPENSIVE_RCS) {
  cat("RC1 skipped (RUN_EXPENSIVE_RCS = FALSE).\n",
      "Prior result: 4 ADM-2 units differ between WGS84 and Mollweide (CRS-insensitive).\n",
      "Set RUN_EXPENSIVE_RCS <- TRUE in constants to re-validate.\n")
} else {

crs_variants <- list(
  WGS84     = st_crs(4326),
  Mollweide = mollweide_crs
)

# Export all objects needed by workers
rc1_results <- future_lapply(crs_variants, function(crs) {
  sf::sf_use_s2(FALSE)

  adm2_proj   <- sf::st_transform(adm2, crs) |> dplyr::select(gid_adm2)
  geoepr_proj <- sf::st_transform(geoepr_data, crs) |> dplyr::select(gwgroupid, from, to)

  links <- sf::st_join(adm2_proj, geoepr_proj,
                       join = sf::st_intersects, left = FALSE) |>
    sf::st_drop_geometry() |>
    dplyr::as_tibble()

  links |>
    dplyr::filter(to >= YEAR_START, from <= YEAR_END) |>
    dplyr::mutate(
      year_from = pmax(as.integer(from), YEAR_START),
      year_to   = pmin(as.integer(to),   YEAR_END),
      year      = purrr::map2(year_from, year_to, seq)
    ) |>
    tidyr::unnest(year) |>
    dplyr::select(gid_adm2, gwgroupid, year) |>
    dplyr::left_join(
      epr_yearly |> dplyr::select(gwgroupid, year, status),
      by = c("gwgroupid", "year")
    ) |>
    dplyr::group_by(gid_adm2, year) |>
    dplyr::summarise(
      EPR_low_bin  = as.integer(any(status %in% low_status,  na.rm = TRUE)),
      EPR_high_bin = as.integer(any(status %in% high_status, na.rm = TRUE)),
      .groups = "drop"
    )
}, future.seed = TRUE,
   future.globals = list(
     adm2 = adm2, geoepr_data = geoepr_data, epr_yearly = epr_yearly,
     YEAR_START = YEAR_START, YEAR_END = YEAR_END,
     low_status = low_status, high_status = high_status
   ))

# Compare for spot-check year
yr_check <- RC1_CHECK_YEAR
rc1_compare <- rc1_results[["WGS84"]] |>
  dplyr::filter(year == yr_check) |>
  dplyr::rename(low_wgs84 = EPR_low_bin, high_wgs84 = EPR_high_bin) |>
  dplyr::left_join(
    rc1_results[["Mollweide"]] |>
      dplyr::filter(year == yr_check) |>
      dplyr::rename(low_moll = EPR_low_bin, high_moll = EPR_high_bin),
    by = "gid_adm2"
  )

cat("RC1 — CRS sensitivity (year", yr_check, "):\n")
cat("  EPR_low_bin  differs:", sum(rc1_compare$low_wgs84  != rc1_compare$low_moll,  na.rm = TRUE), "ADM-2 units\n")
cat("  EPR_high_bin differs:", sum(rc1_compare$high_wgs84 != rc1_compare$high_moll, na.rm = TRUE), "ADM-2 units\n")

} # end if (RUN_EXPENSIVE_RCS)
```

### RC2: Area-weighted overlap threshold

```{r robustness-area-threshold}
# spell_intersections (overlap_pct = int_area / adm2_area) was precomputed
# in Step 4b and is reused here. Just apply different thresholds in parallel.

cat("RC2 reusing spell_intersections from Step 4b (", nrow(spell_intersections), "pairs ).\n")

# Apply thresholds in parallel
thresholds <- c(0, 0.05, 0.10, 0.25)

rc2_results <- future_lapply(thresholds, function(thresh) {
  adm2_group_years_area <- expand_yearly(
      spell_intersections |>
        dplyr::filter(overlap_pct >= thresh) |>
        dplyr::rename(from = spell_from, to = spell_to),
      YEAR_START, YEAR_END
    ) |>
    dplyr::select(gid_adm2, gwgroupid, year)

  adm2_group_years_area |>
    dplyr::left_join(
      epr_yearly |> dplyr::select(gwgroupid, year, status),
      by = c("gwgroupid", "year")
    ) |>
    dplyr::group_by(gid_adm2, year) |>
    dplyr::summarise(
      EPR_low_bin  = as.integer(any(status %in% low_status,  na.rm = TRUE)),
      EPR_high_bin = as.integer(any(status %in% high_status, na.rm = TRUE)),
      .groups = "drop"
    )
}, future.seed = TRUE,
   future.globals = list(
     expand_yearly = expand_yearly,
     spell_intersections = spell_intersections, epr_yearly = epr_yearly,
     YEAR_START = YEAR_START, YEAR_END = YEAR_END,
     low_status = low_status, high_status = high_status
   ))

names(rc2_results) <- paste0("thresh_", thresholds)

# Compare: how many flags change across thresholds?
cat("\nRC2 — area threshold sensitivity (all years):\n")
cat(sprintf("  %-12s  %8s  %9s\n", "Threshold", "n_low=1", "n_high=1"))
for (nm in names(rc2_results)) {
  r <- rc2_results[[nm]]
  cat(sprintf("  %-12s  %8d  %9d\n", nm, sum(r$EPR_low_bin), sum(r$EPR_high_bin)))
}
```

### RC3–RC5: Tabular checks

```{r robustness-temporal-consistency}
cat("\n=== RC3: Temporal consistency ===\n")
temporal_check <- panel |>
  arrange(gid_adm2, year) |>
  group_by(gid_adm2) |>
  summarise(
    n_low_switches  = sum(abs(diff(EPR_low_bin))),
    n_high_switches = sum(abs(diff(EPR_high_bin))),
    .groups = "drop"
  )
cat("EPR_low_bin switches per unit:\n");  print(table(temporal_check$n_low_switches))
cat("EPR_high_bin switches per unit:\n"); print(table(temporal_check$n_high_switches))
cat("Units with >5 switches:", sum(temporal_check$n_low_switches > 5 |
                                    temporal_check$n_high_switches > 5), "\n")
```

```{r robustness-mutual-exclusivity}
cat("\n=== RC4: Coexistence (both flags = 1) ===\n")
n_coexist <- sum(panel$EPR_low_bin == 1 & panel$EPR_high_bin == 1)
cat("ADM-2 × year obs with both low and high:", n_coexist,
    sprintf("(%.2f%% of panel)\n",
            mean(panel$EPR_low_bin == 1 & panel$EPR_high_bin == 1) * 100))
cat("Note: coexistence is expected — different groups in the same ADM-2\n",
    "can simultaneously hold different political power statuses.\n")
```

```{r robustness-coverage}
cat("\n=== RC5: EPR/GeoEPR coverage ===\n")
geoepr_only <- anti_join(geoepr_ids, epr_ids, by = "gwgroupid")
cat("EPR groups:", nrow(epr_ids),
    "| GeoEPR groups:", nrow(geoepr_ids), "\n")
cat("In EPR but not GeoEPR (lost from overlay):", nrow(epr_only), "\n")
cat("In GeoEPR but not EPR (no status):", nrow(geoepr_only), "\n")
```

## Coverage Audit — Phase 2: Full Pipeline Attrition Table

```{r coverage-phase2}
cat("=== COVERAGE AUDIT — Phase 2: Group attrition across the full pipeline ===\n\n")

# Stage 1: All unique EPR groups
n1 <- n_distinct(epr_data$gwgroupid)

# Stage 2: EPR groups with a GeoEPR polygon
n2 <- n_distinct(geoepr_data$gwgroupid)

# Stage 3: GeoEPR groups whose polygon intersects at least one ADM-2 unit.
# adm2_spell_links is built with left = FALSE on the ADM-2 side, so only
# (ADM-2, spell) pairs with a spatial intersection are retained. Groups
# with a polygon that falls entirely outside GADM coverage do not appear.
n3 <- n_distinct(adm2_spell_links$gwgroupid)

# Stage 3 complement: groups with polygon but no ADM-2 intersection
geoepr_no_adm2 <- anti_join(
  geoepr_data |> st_drop_geometry() |> distinct(gwgroupid),
  adm2_spell_links |> distinct(gwgroupid),
  by = "gwgroupid"
)
cat("GeoEPR groups with polygon but no ADM-2 intersection:",
    nrow(geoepr_no_adm2), "\n")

# Stage 4: After temporal filter (2000–2023)
n4 <- adm2_spell_links |>
  filter(to >= YEAR_START, from <= YEAR_END) |>
  pull(gwgroupid) |>
  n_distinct()

# Stage 5: Groups that appear in the panel with a non-NA EPR status
n5 <- adm2_group_years |>
  left_join(epr_yearly |> select(gwgroupid, year, status),
            by = c("gwgroupid", "year")) |>
  filter(!is.na(status)) |>
  pull(gwgroupid) |>
  n_distinct()

attrition <- tibble(
  Stage = c(
    "1. EPR Core — all unique groups",
    "2. With GeoEPR polygon (spatial coverage)",
    "3. Polygon intersects ≥1 ADM-2 unit",
    "4. Active in 2000–2023 (temporal filter)",
    "5. With matched EPR status in panel"
  ),
  N = c(n1, n2, n3, n4, n5)
) |>
  mutate(
    Lost_step    = c(NA_integer_, as.integer(lag(N)[-1] - N[-1])),
    Pct_retained = round(N / first(N) * 100, 1)
  )

print(attrition, n = Inf)
cat("\nNote: 'Lost_step' is loss at each transition; 'Pct_retained' is cumulative from Stage 1.\n")
```

## Step 8: Export

```{r export}
out_path <- here::here("EPR", "output", "adm2_epr_panel.rds")
saveRDS(panel, out_path)
cat("Saved:", out_path, "\n")
cat("Dimensions:", nrow(panel), "rows ×", ncol(panel), "cols\n")
glimpse(panel)

# Shut down parallel workers
plan(sequential)

sessionInfo()
```
