---
title: "ADM-2 EPR Cross-Section"
subtitle: "Time-invariant variables from the EPR panel (2000–2022)"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    theme: cosmo
    code-fold: true
execute:
  warning: false
  message: false
---

## Overview

This script collapses the balanced ADM-2 × year panel (48,010 units × 23 years) into a
**cross-section at the ADM-2 unit level**, computing time-invariant summaries of each
unit's EPR exposure over 2000–2022.

### Output variables

| Variable | Type | Definition |
|----------|------|-----------|
| `always_low` | Binary | `EPR_low_bin == 1` in all 23 years |
| `always_high` | Binary | `EPR_high_bin == 1` in all 23 years |
| `never_low` | Binary | `EPR_low_bin == 0` in all 23 years |
| `never_high` | Binary | `EPR_high_bin == 0` in all 23 years |
| `ever_low` | Binary | `EPR_low_bin == 1` in at least one year |
| `ever_high` | Binary | `EPR_high_bin == 1` in at least one year |
| `mean_low_area_pct` | Continuous [0,1] | Mean of `EPR_low_pct` over 2000–2022 |
| `mean_high_area_pct` | Continuous [0,1] | Mean of `EPR_high_pct` over 2000–2022 |
| `mean_low_pop_pct` | Continuous [0,1] | Mean of `EPR_low_pop_pct` over 2000–2022 |
| `mean_high_pop_pct` | Continuous [0,1] | Mean of `EPR_high_pop_pct` over 2000–2022 |
| `mean_low_country_area_pct` | Continuous [0,1] | Mean of `EPR_low_country_area_pct` over 2000–2022 |
| `mean_high_country_area_pct` | Continuous [0,1] | Mean of `EPR_high_country_area_pct` over 2000–2022 |
| `mean_low_country_pop_pct` | Continuous [0,1] | Mean of `EPR_low_country_pop_pct` over 2000–2022 |
| `mean_high_country_pop_pct` | Continuous [0,1] | Mean of `EPR_high_country_pop_pct` over 2000–2022 |

```{r setup}
if (!require("pacman")) install.packages("pacman", repos = "https://cran.r-project.org")

pacman::p_load(
  here,
  tidyverse,
  kableExtra,
  scales,
  sf,
  patchwork,
  biscale
)

panel <- readRDS(here::here("output", "adm2_epr_panel.rds"))
cat("Panel loaded:", nrow(panel), "rows ×", ncol(panel), "cols\n")
cat("Years:", paste(range(panel$year), collapse = "–"), "\n")
cat("ADM-2 units:", n_distinct(panel$gid_adm2), "\n")

# GADM name lookup
gadm_names <- sf::st_read(
  here::here("source", "gadm_410_adm2_clean.gpkg"),
  quiet = TRUE
) |>
  sf::st_drop_geometry() |>
  dplyr::select(gid_adm2, iso3, country, name_adm2)
```

---

## Step 1: Binary Cross-Section Flags

```{r binary-flags}
n_years <- n_distinct(panel$year)

cross_bin <- panel |>
  group_by(gid_adm2) |>
  summarise(
    always_low  = as.integer(sum(EPR_low_bin)  == n_years),
    always_high = as.integer(sum(EPR_high_bin) == n_years),
    never_low   = as.integer(sum(EPR_low_bin)  == 0),
    never_high  = as.integer(sum(EPR_high_bin) == 0),
    ever_low    = as.integer(any(EPR_low_bin   == 1)),
    ever_high   = as.integer(any(EPR_high_bin  == 1)),
    .groups = "drop"
  )

tibble(
  Variable = c("always_low", "always_high", "never_low", "never_high",
               "ever_low", "ever_high"),
  `n units` = c(sum(cross_bin$always_low), sum(cross_bin$always_high),
                sum(cross_bin$never_low),  sum(cross_bin$never_high),
                sum(cross_bin$ever_low),   sum(cross_bin$ever_high)),
  `% units` = percent(c(mean(cross_bin$always_low), mean(cross_bin$always_high),
                         mean(cross_bin$never_low),  mean(cross_bin$never_high),
                         mean(cross_bin$ever_low),   mean(cross_bin$ever_high)), 0.1)
) |>
  kbl(caption = "Binary cross-section flags across 48,010 ADM-2 units",
      align = c("l", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  footnote(general = "always_* = flag is 1 in all 23 years; never_* = flag is 0 in all 23 years; ever_* = flag is 1 in at least one year.",
           general_title = "Note:")
```

---

## Step 2: Continuous Averages

```{r continuous-means}
cross_cont <- panel |>
  group_by(gid_adm2) |>
  summarise(
    mean_low_area_pct          = mean(EPR_low_pct),
    mean_high_area_pct         = mean(EPR_high_pct),
    mean_low_pop_pct           = mean(EPR_low_pop_pct),
    mean_high_pop_pct          = mean(EPR_high_pop_pct),
    mean_low_country_area_pct  = mean(EPR_low_country_area_pct),
    mean_high_country_area_pct = mean(EPR_high_country_area_pct),
    mean_low_country_pop_pct   = mean(EPR_low_country_pop_pct),
    mean_high_country_pop_pct  = mean(EPR_high_country_pop_pct),
    .groups = "drop"
  )

tibble(
  Variable = names(cross_cont)[-1],
  Mean     = sapply(cross_cont[-1], mean)   |> round(5),
  Median   = sapply(cross_cont[-1], median) |> round(5),
  SD       = sapply(cross_cont[-1], sd)     |> round(5),
  `% zero` = sapply(cross_cont[-1], \(x) percent(mean(x == 0), 0.1))
) |>
  kbl(caption = "Summary statistics for continuous cross-section variables",
      align = c("l", "r", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  pack_rows("ADM-relative (area)",       1, 2) |>
  pack_rows("ADM-relative (population)", 3, 4) |>
  pack_rows("Country-relative (area)",    5, 6) |>
  pack_rows("Country-relative (population)", 7, 8) |>
  footnote(general = "Means are time-averages over 23 years (2000–2022) per ADM-2 unit, then summarised across all 48,010 units. % zero = share of units with a zero time-average (no EPR group presence in any year).",
           general_title = "Note:")
```

---

## Step 3: Assemble Cross-Section Dataset

```{r assemble}
crosssection <- cross_bin |>
  left_join(cross_cont, by = "gid_adm2") |>
  left_join(gadm_names,  by = "gid_adm2")

cat("Cross-section dimensions:", nrow(crosssection), "units ×",
    ncol(crosssection), "variables\n")
cat("Columns:", paste(names(crosssection), collapse = ", "), "\n")
```

---

## Step 4: Diagnostics

```{r diag-binary}
# Mutual exclusivity checks
cat("=== Binary flag consistency ===\n")
cat("always_low  & never_low  (impossible):",
    sum(crosssection$always_low  & crosssection$never_low),  "\n")
cat("always_high & never_high (impossible):",
    sum(crosssection$always_high & crosssection$never_high), "\n")
cat("always_low  implies ever_low  (must be 0 violations):",
    sum(crosssection$always_low  & !crosssection$ever_low),  "\n")
cat("always_high implies ever_high (must be 0 violations):",
    sum(crosssection$always_high & !crosssection$ever_high), "\n")
```

```{r diag-continuous}
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Distribution of mean unit-relative population share per ADM-2 unit (2000–2022). Left: share of units with exactly zero EPR overlap. Right: conditional distribution among non-zero units (log scale)."

diag_df <- crosssection |>
  select(mean_low_pop_pct, mean_high_pop_pct) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(variable = recode(variable,
    mean_low_pop_pct  = "Low-power",
    mean_high_pop_pct = "High-power"
  ))

# Panel A: zero-mass bar
p_zero <- diag_df |>
  group_by(variable) |>
  summarise(pct_zero = mean(value == 0), .groups = "drop") |>
  ggplot(aes(x = variable, y = pct_zero, fill = variable)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = percent(pct_zero, accuracy = 0.1)),
            vjust = -0.4, size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = percent, limits = c(0, 1),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = "Share of ADM-2 units", title = "A. Share with zero overlap") +
  theme_minimal(base_size = 12)

# Panel B: non-zero density (log x-axis)
p_nonzero <- diag_df |>
  filter(value > 0) |>
  ggplot(aes(x = value, colour = variable, fill = variable)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  scale_x_log10(labels = percent) +
  scale_colour_brewer(palette = "Set2", name = NULL) +
  scale_fill_brewer(palette = "Set2", name = NULL) +
  labs(x = "Mean unit-level population share (log scale)", y = "Density",
       title = "B. Among non-zero units") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

patchwork::wrap_plots(p_zero, p_nonzero, widths = c(1, 2))
```

### Variable selection for cross-sectional analysis

For subnational aid allocation studies using EPR as a proxy for political power, the
recommended variable is **`mean_low_pop_pct`**, the unit-relative population share.
It measures the average share of the ADM-2 unit's *own* population living inside
low-power group territory over 2000–2022, capturing local exclusion intensity as a
donor observer would perceive it on the ground.

**Why unit-relative (`mean_low_pop_pct`) over country-relative (`mean_low_country_pop_pct`):**

- `mean_low_pop_pct` answers: *how much of this district's population is politically
  marginalized?* This is the relevant margin for local aid targeting decisions.
- `mean_low_country_pop_pct` answers: *what share of the country's total population
  does this district's marginalized population represent?* This is a national salience
  measure, better suited for constructing country-level aggregates or instruments.
- Country-relative values are mechanically tiny (ADM-2 pop / country pop is typically
  < 0.01), compressing variation and producing near-zero distributions even within
  EPR-coded countries.

**Recommended specification:**

| Variable | Role | Rationale |
|---|---|---|
| `mean_low_pop_pct` | Primary treatment | Intensity of low-power exposure in the unit |
| `mean_high_pop_pct` | Control / heterogeneity | Dominant-group presence may independently predict aid |
| `ever_low` | Binary alternative | Clean treated/control distinction if intensity not needed |

```{r map-bivariate}
#| fig-width: 11
#| fig-height: 6
#| fig-cap: "Bivariate choropleth: joint distribution of mean low-power and high-power unit-relative population shares at the ADM-2 level (2000–2022 average). Blue: high share of ADM-2 population inside low-power territory. Dark: high share inside high-power territory. Dark blue: coexistence. Light grey: no EPR group presence. Jenks natural breaks (dim = 3) per variable."

# Load GADM ADM-2 geometry (geometry retained for mapping)
adm2_sf <- sf::st_read(
  here::here("source", "gadm_410_adm2_clean.gpkg"), quiet = TRUE
) |>
  dplyr::select(gid_adm2)

# Join crosssection attributes to geometry.
# Use unit-relative population shares (mean_low_pop_pct, mean_high_pop_pct):
# these capture local exclusion intensity and vary meaningfully across units.
# Country-relative shares (mean_low_country_pop_pct) are near-zero by construction
# and add little visual information in a map.
map_df <- adm2_sf |>
  dplyr::left_join(
    crosssection |>
      dplyr::select(gid_adm2, mean_low_pop_pct, mean_high_pop_pct),
    by = "gid_adm2"
  )

# Assign bivariate classes (3×3 grid per variable).
# style = "jenks" (natural breaks) handles zero-inflated distributions:
# quantile breaks fail here because most values are 0, causing non-unique cuts.
map_biv <- biscale::bi_class(
  map_df,
  x     = mean_low_pop_pct,
  y     = mean_high_pop_pct,
  style = "jenks",
  dim   = 3
)

# Map panel
p_map <- ggplot(map_biv) +
  geom_sf(aes(fill = bi_class), color = NA, show.legend = FALSE) +
  biscale::bi_scale_fill(pal = "DkBlue", dim = 3) +
  labs(
    title    = "Low-power vs high-power group presence (bivariate)",
    subtitle = "Mean share of ADM-2 population inside group territory, 2000–2022 average"
  ) +
  theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# Legend panel
p_legend <- biscale::bi_legend(
  pal  = "DkBlue",
  dim  = 3,
  xlab = "Low-power ↑",
  ylab = "High-power ↑",
  size = 9
)

# Combine: wide map + compact legend
patchwork::wrap_plots(p_map, p_legend, widths = c(4, 1))
```

---

## Step 5: Export

```{r export}
out_path <- here::here("output", "adm2_epr_crosssection.rds")
saveRDS(crosssection, out_path)
cat("Saved:", out_path, "\n")
cat("Dimensions:", nrow(crosssection), "rows ×", ncol(crosssection), "cols\n")

sessionInfo()
```
